\documentclass[11pt,a4paper]{article}

\usepackage{fontspec}
\setmainfont{Palatino}
\usepackage{geometry}
\geometry{margin=2.5cm}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=black, citecolor=black, urlcolor=blue}

% Convenience macros
\newcommand{\Os}{\Omega^{*}}
\newcommand{\G}{\mathrm{G}}
\newcommand{\f}{f}

\sloppy

\title{A Two-Stage Perceptual Account of Interval Consonance}
\author{David De Roure}
\date{Working note, June 2026}

\begin{document}
\maketitle

% ─── Abstract ────────────────────────────────────────────────────────────────

\begin{abstract}
Euler's \emph{Gradus Suavitatis} (1739) assigns a dissonance value to a musical
interval $p/q$ by the formula $\G(p/q) = 1 + \Os(p) + \Os(q)$, where
$\Os(n) = \sum_i e_i(p_i - 1)$ sums the weighted prime exponents of~$n$.
We show that the simpler asymmetric formula $f(p/q) = p + \Os(q)$ achieves
a higher rank correlation with human consonance ratings ($\rho = 0.989$
vs.\ $0.979$), and that this formula admits a natural two-stage perceptual
interpretation grounded in harmonic series theory. We also show that Gradus
is exactly equivalent to a weighted harmonic coincidence count with weights
$w(n) = \Os(n)$, and that the simpler metric $\max(p,q)$ --- equivalent to
Galileo's pulse-coincidence model --- achieves the same correlation as
$f(p/q)$.
\end{abstract}

% ─── 1. Background ───────────────────────────────────────────────────────────

\section{Background}

Euler's Gradus Suavitatis, catalogued in OEIS A275314, assigns to each musical
interval a positive integer measuring its arithmetic dissonance. For a ratio
$p/q$ in lowest terms, the characteristic number is $n = p \cdot q$, and:
\[
  \G(p/q) = 1 + \Os(p \cdot q)
\]
where $\Os$ is the completely additive function $\Os(n) = \sum_i e_i(p_i - 1)$.
Since $\Os$ is completely additive, $\G(p/q) = 1 + \Os(p) + \Os(q)$, treating
numerator and denominator symmetrically. The weights $(p_i - 1)$ for prime
$p_i$ encode Euler's aesthetic judgment that large primes represent greater
``difficulty.''

Euler's formula correctly ranks the 13 standard intervals of Western music
against human consonance ratings \cite{krumhansl1990} with Spearman
$\rho = 0.979$, and is exactly tied only on the major third/major sixth pair
and on a three-way tie at Gradus~8.

% ─── 2. Gradus as harmonic coincidence ───────────────────────────────────────

\section{Gradus as Harmonic Coincidence}

For two notes at frequency ratio $p:q$ (coprime), exact harmonic coincidences
occur at harmonics $p, 2p, 3p, \ldots$ of the lower note. For a fixed count
$M$ of coincidences, the weighted sum with weights $w(n) = \Os(n)$ yields:
\[
  \sum_{k=1}^{M} \Os(k \cdot p) \;=\; \Os(p) \cdot M + C(M)
\]
where $C(M) = \sum_{k=1}^{M} \Os(k)$ is a constant independent of~$p$. This
follows from the complete additivity of $\Os$: $\Os(k \cdot p) = \Os(k) +
\Os(p)$. The symmetric score over both notes gives:
\[
  \text{Score} = M \cdot (\G(p/q) - 1) + 2 \, C(M)
\]

\textbf{Gradus is therefore exactly a weighted harmonic coincidence count},
with harmonics weighted by their own prime complexity. This confirms the
physical intuition behind Euler's formula but reveals its self-referential
structure: recovering Gradus from coincidences requires weights that are
themselves Gradus-like.

Unweighted coincidence counting (flat weights) achieves $\rho = 0.791$ across
all coprime ratios with $p, q \leq 16$. The weight function $\log n$ achieves
$\rho = 0.795$ without any prime structure, establishing a ceiling for
physics-based models that do not invoke prime factorisation.

% ─── 3. Galileo and max(p,q) ─────────────────────────────────────────────────

\section{Galileo's Model and \texorpdfstring{$\max(p,q)$}{max(p,q)}}

Galileo (1638) described consonance in terms of pulse coincidences: two strings
at ratio $p:q$ produce coinciding pulses every $p$ vibrations of the lower
string. The ``sweetness'' of the interval grows with the frequency of
coincidences, i.e.\ inversely with~$p$. The metric $\max(p,q) = p$ is
therefore:

\begin{enumerate}
  \item \textbf{Galileo's pulse model} --- vibrations between coincidences;
  \item \textbf{Harmonic series position} --- the upper note is partial $p$ of
        the implied fundamental \cite{rameau1722};
  \item \textbf{Farey index} --- the smallest $n$ such that $q/p$ appears in
        the Farey sequence $F_n$ (Cauchy 1816);
  \item \textbf{First coincidence harmonic} --- the position in the lower
        note's series where the two notes first meet.
\end{enumerate}

These four descriptions, discovered independently across four centuries, are
mathematically identical. $\max(p,q)$ achieves $\rho = 0.989$ against human
ratings, marginally exceeding Gradus, but cannot distinguish intervals with
the same numerator (e.g.\ $5/4$ and $5/3$ both have $\max = 5$).

% ─── 4. An improved formula ──────────────────────────────────────────────────

\section{An Improved Formula}

Euler's formula is symmetric: $\Os(p) + \Os(q)$. Human ratings, however,
distinguish numerator and denominator --- the bass note and the upper note
play different perceptual roles. We find empirically that the asymmetric
formula:
\[
  \boxed{f(p/q) = p + \Os(q)}
\]
achieves $\rho = 0.989$, matching $\max(p,q)$ and exceeding Gradus. The
formula differs from Gradus only in replacing $\Os(p)$ with~$p$ for the
numerator. Optimising prime weights freely with six parameters (separate
weights for primes 2, 3, 5 in numerator and denominator) achieves perfect
rank correlation $\rho = 1.000$, confirming that the structure required to
perfectly predict the human rankings is present in an asymmetric
prime-weighted formula.

\medskip
\noindent The formula $f(p/q) = p + \Os(q)$ decomposes as the sum of two
quantities:

\medskip
\begin{tabular}{lll}
\toprule
Component & Formula & Meaning \\
\midrule
Upper note reach & $p$ & Partial number of upper note in implied harmonic series \\
Bass context depth & $\Os(q)$ & Prime complexity of lower note's position in series \\
\bottomrule
\end{tabular}

% ─── 5. Two-stage perceptual model ───────────────────────────────────────────

\section{A Two-Stage Perceptual Model}

The formula motivates a two-stage model of how the auditory system evaluates
an interval:

\medskip\noindent
\textbf{Stage 1 --- Bass establishes harmonic context (cost $\Os(q)$).}
The lower note is the $q$-th partial of an implied fundamental $F$. The brain
extrapolates downward to $F$ by ``dividing out'' the prime factors of~$q$.
Each factor of 2 (an octave) costs 1 unit; each factor of 3 costs~2; each
factor of 5 costs~4. This is exactly $\Os(q)$. Intervals whose bass note is a
power of 2 (e.g.\ the octave, $q=1$; the major second bass, $q=8=2^3$)
establish the fundamental cheaply via octave equivalence. Intervals with an
odd prime in $q$ (the fourth, $q=3$; the minor third bass, $q=5$) require the
brain to extrapolate through a genuinely new prime, incurring higher cost.

\medskip\noindent
\textbf{Stage 2 --- Upper note reaches into the series (cost $p$).}
Once the fundamental is established, the upper note must be recognised as its
$p$-th partial. Higher partials are quieter (amplitude $\propto 1/p$ in
typical harmonic sounds), more narrowly tuned, more sensitive to mistuning,
and have lower pitch salience \cite{terhardt1979}. The cost of confirming the
upper note as a member of the harmonic series is therefore $p$ itself.

\medskip\noindent
\textbf{Total perceptual effort:} $f(p/q) = p + \Os(q)$.

\medskip\noindent
This account connects:
\begin{itemize}
  \item \textbf{Galileo} (1638): pulse coincidences as physical mechanism;
  \item \textbf{Rameau} (1722): the \emph{corps sonore} (harmonic series) as
        the source of musical meaning;
  \item \textbf{Euler} (1739): prime arithmetic as the language of harmonic
        complexity;
  \item \textbf{Terhardt} (1979): virtual pitch and harmonic template matching
        as auditory mechanisms;
  \item \textbf{Huron} (2001): statistical learning of harmonic patterns as
        the basis of consonance judgments.
\end{itemize}

% ─── 6. The remaining failure ─────────────────────────────────────────────────

\section{The Remaining Failure}

No simple formula distinguishes the major third ($5/4$, human rank~5) from
the major sixth ($5/3$, human rank~7). Both have $p = 5$ and $\Os(q) = 2$
(since $\Os(4) = \Os(2^2) = 2$ and $\Os(3) = 2$). The distinction requires
knowing that $q = 4 = 2^2$ is simpler than $q = 3$ under octave equivalence,
yet the prime-weighting function cannot see this.

Lahdelma \& Eerola \cite{lahdelma2016} decompose consonance ratings into
sensory and familiarity components; on \emph{sensory consonance} alone,
$5/4$ and $5/3$ are nearly indistinguishable across listeners, suggesting
that the tie in $f(p/q)$ may accurately reflect the purely arithmetic
component of the percept. The gap observed in Krumhansl's data is plausibly
attributable to learned exposure --- major thirds occur far more frequently
than major sixths in tonal music --- consistent with the finding of
McDermott et al.\ \cite{mcdermott2016} that consonance preferences in a
culture with minimal Western music exposure are substantially reduced.
Resolving the tie within an arithmetic formula would therefore require
modelling cultural familiarity, which is outside the scope of the present
approach.

% ─── 7. Summary ──────────────────────────────────────────────────────────────

\section{Summary}

\medskip
\begin{tabular}{lll}
\toprule
Formula & $\rho$ vs.\ human & Notes \\
\midrule
Euler's Gradus: $1 + \Os(p) + \Os(q)$ & 0.979 & Symmetric; two ties unresolved \\
Galileo / $\max(p,q)$                  & 0.989 & Simplest; same tie \\
$f(p/q) = p + \Os(q)$                  & \textbf{0.989} & Asymmetric; two-stage interpretation \\
Optimised asymmetric weights            & 1.000 & Six parameters; no closed form \\
Roughness (Plomp--Levelt)              & 0.709--0.863 & Register-dependent; fails on tritone$^*$ \\
\bottomrule
\end{tabular}

\medskip
\noindent$^*$ Helmholtz \cite{helmholtz1863} introduced the beating-based
dissonance curve that Plomp \& Levelt \cite{plomplevelt1965} later parametrised
from listener data; the two models produce equivalent rankings for our
purposes.

\bigskip
\noindent Table~\ref{tab:intervals} shows all three formulae alongside the
human consonance rank for the 13 standard just intervals of Western music,
ordered by human rating. Ties in $f(p/q)$ and in $\max(p,q)$ are indicated
by matching values; the sole unresolved tie (major third / major sixth, both
$f = 7$) is discussed in Section~6.

\begin{table}[h]
\centering
\caption{Dissonance values for the 13 standard just intervals, ordered by
human consonance rank \cite{krumhansl1990} (rank 1 = most consonant).}
\label{tab:intervals}
\smallskip
\begin{tabular}{llrrrr}
\toprule
Interval & Ratio & $\G(p/q)$ & $f(p/q)$ & $\max(p,q)$ & Human rank \\
\midrule
Unison        & $1/1$   &  1 &  1 &  1 &  1 \\
Octave        & $2/1$   &  2 &  2 &  2 &  2 \\
Fifth         & $3/2$   &  4 &  4 &  3 &  3 \\
Fourth        & $4/3$   &  5 &  6 &  4 &  4 \\
Major third   & $5/4$   &  7 &  7 &  5 &  5 \\
Minor third   & $6/5$   &  8 & 10 &  6 &  6 \\
Major sixth   & $5/3$   &  7 &  7 &  5 &  7 \\
Minor sixth   & $8/5$   &  8 & 12 &  8 &  8 \\
Major second  & $9/8$   &  8 & 12 &  9 &  9 \\
Minor seventh & $9/5$   &  9 & 13 &  9 & 10 \\
Major seventh & $15/8$  & 10 & 18 & 15 & 11 \\
Minor second  & $16/15$ & 11 & 22 & 16 & 12 \\
Tritone       & $45/32$ & 14 & 50 & 45 & 13 \\
\bottomrule
\end{tabular}
\end{table}

% ─── 8. Comparison ───────────────────────────────────────────────────────────

\section{Comparison of the Three Formulae}

Table~\ref{tab:intervals} invites a direct comparison of the three arithmetic
formulae. Several observations follow.

\paragraph{Agreement on the extremes.}
All three formulae agree without qualification on the most and least consonant
intervals. The unison (1), octave (2), fifth (3/2), and fourth (4/3) are
assigned the lowest values by every measure, consistent with their historical
priority in counterpoint and their prominence in non-Western musical systems.
The tritone (45/32) receives the highest value under every formula, and the
minor and major seconds cluster near the top of the dissonance scale in all
three cases. The disagreements are concentrated in the middle of the table,
among intervals that are also the most contested in historical music theory.

\paragraph{Correlation and ties.}
Euler's Gradus achieves a Spearman rank correlation of $\rho = 0.979$ against
the human data; $f$ and $\max$ both achieve $\rho = 0.989$. The difference is
modest, and caution is warranted given the small sample (13 intervals) and the
cultural specificity of the rating data (see Section~6). More informative than
the correlation coefficient is the tie structure. Gradus produces two tie
groups among the 13 intervals: the major third and major sixth share $G = 7$,
and the minor third, minor sixth, and major second share $G = 8$. The
three-way tie at $G = 8$ is particularly awkward, as the three intervals span
three consecutive human ranks (6, 8, 9). The formula $f$ resolves this
three-way tie --- giving $f = 10$, $12$, $12$ respectively --- though it
introduces a two-way tie between the minor sixth and major second (both $f =
12$, human ranks 8 and 9). The metric $\max(p,q)$ shares the major third /
major sixth tie with $f$, but adds a further tie between the major second and
minor seventh (both $\max = 9$, human ranks 9 and 10). On the criterion of
tie-resolution among these 13 intervals, $f$ performs best, Gradus
intermediate, and $\max$ least well.

\paragraph{Dynamic range.}
The three formulae differ most sharply in their treatment of the tritone. Gradus
gives $G(45/32) = 14$, only four units above the major seventh ($G = 10$).
Both $f$ and $\max$ assign substantially higher values ($f = 50$, $\max = 45$),
placing the tritone in a category of its own and reflecting the large
numerator $p = 45$ that the asymmetric formulae weight directly. Whether this
wider dynamic range better represents human experience is difficult to assess
from rank-correlation data alone; what can be said is that the tritone is
consistently rated as by far the most dissonant interval in the human data,
and both $f$ and $\max$ capture that separation more emphatically than Gradus.

\paragraph{Symmetry.}
Gradus is symmetric in $p$ and $q$: $G(p/q) = G(q/p)$. This reflects Euler's
focus on the product $p \cdot q$ as the relevant arithmetic quantity, and is
appropriate if the perceptual roles of the two notes are interchangeable.
Both $f$ and $\max$ are asymmetric, treating the numerator (upper note) and
denominator (lower note) differently, as is the case when one note functions
as a bass and the other as a melody note. The symmetry of Gradus means it
cannot distinguish $9/5$ from $16/9$ --- two minor sevenths that differ by
only 1 cent but arise from different tuning systems --- whereas $f$ gives
$f(9/5) = 13$ and $f(16/9) = 20$. Whether this asymmetry reflects a genuine
perceptual distinction or an artefact of the formula is an open question,
though the perceptual account in Section~5 offers a principled motivation.

\paragraph{Simplicity and interpretability.}
The metric $\max(p,q)$ is strikingly simple, requires no prime factorisation,
and admits four independent mathematical characterisations (Section~3). These
properties make it an attractive baseline. Its weakness is precisely that
simplicity: it is insensitive to whether the denominator is a power of 2 or
a large prime, treating $4/3$ and $7/3$ identically in the denominator.
Gradus captures this prime sensitivity through $\Os(q)$ but applies the same
weighting to the numerator, where the evidence suggests a different treatment
is warranted. The formula $f$ retains prime sensitivity in the denominator
while replacing the numerator term with the raw partial number, motivated by
the two-stage perceptual account. Each formula thus represents a different
theory of what makes an interval consonant, and the empirical data do not
decisively separate them.

\paragraph{Choice of formula.}
The three formulae are more complementary than competing. For a context
requiring only a rough ordering of standard intervals, any of the three
suffices. Where fine distinctions matter --- competing tuning systems,
intervals outside the standard 13, or analysis of the consonance landscape
under equal temperament --- the formulae diverge in informative ways, and the
choice should be guided by the theoretical question at hand.

% ─── 9. Landscape ────────────────────────────────────────────────────────────

\section{The Consonance Landscape and Equal Temperament}

The three formulae are defined on rational numbers, but musical intervals in
practice are never exact. To study how each formula behaves near a just ratio
we sweep the octave in 1-cent steps, find the simplest rational approximation
$p/q$ (denominator $\leq 64$) within $\pm 15$ cents of each position, and
evaluate $G$, $f$, and $\max$. The result is a \emph{consonance landscape}
--- a spiky function of pitch interval in which low values cluster sharply
around simple fractions and rise steeply on either side.

\subsection*{Well depth and sharpness}

Table~\ref{tab:rivals} shows, for each standard just interval, the nearest
rival rational within $\pm 50$ cents and the corresponding jump in each
formula. The wells around simple fractions are deep for all three functions,
but $f(p/q)$ is substantially steeper: moving 2--5 cents off the fifth (3/2)
takes $f$ from 4 to 104 ($+100$), compared with Gradus rising from 4 to 58
($+54$). The major third is even more extreme: 3 cents off 5/4, $f$ leaps
from 7 to 136, while Gradus rises from 7 to 81. This sharper gradient means
$f(p/q)$ predicts tighter perceptual boundaries around just intonation than
Gradus does --- a potentially testable difference.

\begin{table}[h]
\centering
\caption{Nearest rival rational for each just interval and the resulting jump
in each formula. ``Rival'' is the simplest fraction within $\pm 50$ cents
that is distinct from the just ratio; error is the rival's distance in cents.}
\label{tab:rivals}
\smallskip
\begin{tabular}{llrrrrrr}
\toprule
Interval & Just & Rival & Err (¢) & $\Delta G$ & $\Delta f$ & $\Delta \max$ \\
\midrule
Major second  & $9/8$   & $64/57$  & 2 & $+19$ & $+72$  & $+55$ \\
Minor third   & $6/5$   & $77/64$  & 2 & $+15$ & $+73$  & $+71$ \\
Major third   & $5/4$   & $76/61$  & 3 & $+74$ & $+129$ & $+71$ \\
Fourth        & $4/3$   & $83/62$  & 4 & $+109$& $+108$ & $+79$ \\
Tritone       & $45/32$ & $52/37$  & 1 & $+37$ & $+38$  & $+7$  \\
Fifth         & $3/2$   & $94/63$  & 5 & $+54$ & $+100$ & $+91$ \\
Minor sixth   & $8/5$   & $101/63$ & 2 & $+103$& $+99$  & $+93$ \\
Major sixth   & $5/3$   & $103/62$ & 3 & $+127$& $+127$ & $+98$ \\
Minor seventh & $9/5$   & $16/9$   & 1 & $0$   & $+7$   & $+7$  \\
Major seventh & $15/8$  & $118/63$ & 1 & $+60$ & $+110$ & $+103$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection*{Equal temperament}

Twelve-tone equal temperament (12-TET) divides the octave into 12 equal
semitones of 100 cents each; no interval except the unison and octave is
exactly just. Table~\ref{tab:tet} gives the nearest just approximation for
each 12-TET semitone and the formula values at that approximation.

\begin{table}[h]
\centering
\caption{12-TET semitones: nearest just rational (denominator $\leq 64$,
tolerance $\pm 15$ cents) and formula values.}
\label{tab:tet}
\smallskip
\begin{tabular}{lrllrrr}
\toprule
Note & Cents & Just ratio & Error (¢) & $G$ & $f$ & $\max$ \\
\midrule
C  (unison)  &    0 & $1/1$   & 0.00 &   1 &   1 &   1 \\
C\#          &  100 & $53/50$ & 0.88 &  62 &  62 &  53 \\
D            &  200 & $55/49$ & 0.02 &  27 &  67 &  55 \\
E$\flat$     &  300 & $44/37$ & 0.03 &  49 &  80 &  44 \\
E            &  400 & $63/50$ & 0.11 &  20 &  72 &  63 \\
F  (fourth)  &  500 & $4/3$   & 1.96 &   5 &   6 &   4 \\
F\#          &  600 & $58/41$ & 0.51 &  70 &  98 &  58 \\
G  (fifth)   &  700 & $3/2$   & 1.96 &   4 &   4 &   3 \\
A$\flat$     &  800 & $100/63$& 0.11 &  21 & 110 & 100 \\
A            &  900 & $37/22$ & 0.03 &  48 &  48 &  37 \\
B$\flat$     & 1000 & $98/55$ & 0.02 &  28 & 112 &  98 \\
B            & 1100 & $117/62$& 0.60 &  48 & 148 & 117 \\
C' (octave)  & 1200 & $2/1$   & 0.00 &   2 &   2 &   2 \\
\bottomrule
\end{tabular}
\end{table}

\noindent
Only four of the twelve semitones land near a simple just ratio: the unison,
fourth, fifth, and octave. The remaining eight are approximated by fractions
with large numerators and denominators, giving uniformly high dissonance
values under all three formulae. The consonance of, for example, the
equal-tempered major third (400 cents, nearest ratio $63/50$, $f = 72$) thus
depends entirely on the ear accepting a 14-cent deviation from the just $5/4$
--- pulling the interval into the low-$f$ well of that simple fraction. Equal
temperament is, in this sense, a system of controlled approximation: it
exploits perceptual tolerance around a small number of simple ratios while
treating all other intervals as effectively irrational.

\subsection*{The minor seventh and competing tunings}

The minor seventh provides a striking special case. The Ptolemaic minor
seventh $9/5$ (996 cents) and the Pythagorean minor seventh $16/9$ (995
cents) are only 1 cent apart, yet have distinct arithmetic characters.
Gradus gives both the same value ($G = 9$), and is therefore unable to
distinguish these historically competing tunings. The formula $f$ does
distinguish them: $f(9/5) = 13$ vs $f(16/9) = 20$, favouring the Ptolemaic
tuning. The Pythagorean version $16/9 = (4/3)^2$ arises as two stacked fourths and
is the minor seventh generated by the Pythagorean system, in which all
intervals derive from the perfect fifth. Its higher $f$ value reflects its
larger numerator: the upper note sits at the 16th partial, far higher in the
harmonic series than the 9th partial of the Ptolemaic version. Gradus, being
symmetric in $p$ and $q$, cannot see this distinction ($\Os(16) + \Os(9) =
\Os(9) + \Os(16)$); the asymmetry of $f$ is what exposes it.

% ─── References ──────────────────────────────────────────────────────────────

\begin{thebibliography}{99}

\bibitem{euler1739}
L.~Euler,
\emph{Tentamen novae theoriae musicae},
St.\ Petersburg, 1739.

\bibitem{galilei1638}
G.~Galilei,
\emph{Discorsi e Dimostrazioni Matematiche},
Leiden, 1638.

\bibitem{helmholtz1863}
H.~von Helmholtz,
\emph{On the Sensations of Tone as a Physiological Basis for the Theory of Music},
Vieweg, Braunschweig, 1863.
(English translation by A.~J.~Ellis, Longmans, Green \& Co., London, 1875.)

\bibitem{rameau1722}
J.-P.~Rameau,
\emph{Trait\'{e} de l'Harmonie},
Paris, 1722.

\bibitem{plomplevelt1965}
R.~Plomp and W.~J.~M.~Levelt,
Tonal consonance and critical bandwidth,
\emph{Journal of the Acoustical Society of America} \textbf{38} (1965), 548--560.

\bibitem{terhardt1979}
E.~Terhardt,
Calculating virtual pitch,
\emph{Hearing Research} \textbf{1} (1979), 155--182.

\bibitem{krumhansl1990}
C.~L.~Krumhansl,
\emph{Cognitive Foundations of Musical Pitch},
Oxford University Press, 1990.

\bibitem{lahdelma2016}
I.~Lahdelma and T.~Eerola,
Single chords convey distinct emotional qualities to both na\"{i}ve and
experienced listeners,
\emph{Psychology of Music} \textbf{44} (2016), 37--54.

\bibitem{mcdermott2016}
J.~H.~McDermott, A.~F.~Schultz, E.~A.~Undurraga and R.~A.~Godoy,
Indifference to dissonance in native Amazonian,
\emph{Nature} \textbf{535} (2016), 547--550.

\bibitem{huron2001}
D.~Huron,
Tone and voice: a derivation of the rules of voice-leading from perceptual
principles,
\emph{Music Perception} \textbf{19} (2001), 1--64.

\bibitem{bowling2018}
D.~L.~Bowling, D.~Purves and K.~Z.~Gill,
Vocal similarity predicts the relative attraction of musical chords,
\emph{Proceedings of the National Academy of Sciences} \textbf{115} (2018),
216--221.

\end{thebibliography}

\end{document}
